---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
set.seed(42) #setting seed for readability
population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sample_100 <- population |> 
  sample_n(100)
sample_50 <- population |> 
  sample_n(50)
sample_10 <- population |> 
  sample_n(10)
```


### Task 4: Calculate the mean age and standard deviation of the sample
```{r}
sample_100 |> summarize(mean = mean(age), sd = sd(age))
sample_50 |> summarize(mean = mean(age), sd = sd(age))
sample_10 |> summarize(mean = mean(age), sd = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population
```{r}
population |> summarize(mean = mean(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

My sample mean and standard deviation was 48.8 and 18.9 respectively and for my population 49.9 and 17.5. These are pretty similar but as I decreased my sample size my data got less accurate. With my sample size of 10 my mean was 41.2 which is a completely different result in comparison to the entire population. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
