---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```



### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv
```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```


### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`
```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024` - `Fall 2021`)/`Fall 2021` * 100)
```



### Task 4: Describe our results

Students paid $267 more in fees than in the Fall of 2024 than the Fall of 2021 with the Student Sustainability fee having one of the biggest increases at 150%. The Student Counseling Center fee was the only category to surpass it because students paid nothing in the Fall of 2021 and were up to $50 by the Fall of 2024. The Athletics' fee was the only category to remain the same. My story angle would be trying to uncover why the fees have increased so much.

Questions:
- What is the difference between Athletics and the other categories? What caused this fee to remain the same as the remaining categories increased?
- What changes have been made within the Student Counseling Center between Fall 2021 and Fall 2024?
- The Student Activities' fee saw the most minimal change, does this mean the activities have not changed as much over the years or have the changes been funded by people other than the students?

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
