---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "YOUR NAME"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

This is the opposite of the class activity because this scatter plot depicts a strong negative relationship whereas the activity showoed a weak positive one.


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

p-value: 0.00000000000000022, the Adjusted R-squared value: .673, and the coefficients: intercept: 68.049, farms pct: -0.595

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

This model shows a strong and clear negative relationship between FARMS percentage and math scores so as the percentage of students on free and reduced meals goes up, math scores tend to go down. The p-value is extremely low and the adjusted R-squared is high, so the model is statistically solid and explains a lot of the score differences. The model is definitely worth exploring further to better understand how poverty may be affecting student performance.

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

Both Fox Chapel Elementary and Broad Acres Elementary have the highest residuals and performed significantly better than anticipated. The next two highest, Germantown and Somerset performed significantly lower than expected. Fox Chapel Elementary (20876) is pretty educated and affluent area, 51.6% of their population has a Bachelor's degree or higher. After seeing this I would expect them to perform better because they are not considered low-income. Broad Acres Elementary (20903) is not as wealthy or educated. I think it's most significant that around 34% of residents haven't completed hs and an even lower percentage received a Bachelor's degree, this made their performance a little more surprising for me but they still are not considered to be low-income so I would noot expect the students to score badly.

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

The model shows that schools with more low-income students tend to have lower math scores. For every 1% increase in students needing free or reduced meals, math performance drops by about 0.6 percentage points. This suggests a strong link between poverty and lower achievement in math. A story pitch could focus on how a school board member plans to address these disparities and support students in high-poverty schools.

