---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: Nyla Cherry
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

The correlation matrix helps show which factors move together with voter turnout. Income and education seem to be positively related to turnout—counties with higher income and education tend to vote more. Population size and unemployment don’t show a clear pattern, meaning they may not be strongly related to turnout.


```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

The chart shows that some things, like higher income, older age, and more education, are linked to more people voting. Other things, like higher unemployment and bigger population, are linked to fewer people voting. The strength of these relationships varies, and some relationships seem stronger than others like the unemployed percentage, but we can't fully judge significance without statistical values like p-values or R-squared. Overall, the model gives some useful insight, but without those detailed metrics, we can't be sure how well it predicts turnout or if all variables are truly important.

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

The chart shows that Model 2, Model 4, and the Full model all have very similar adjusted R-squared values, meaning they explain about the same amount of variation in voter turnout. Model 3 has a lower value, so it doesn’t predict turnout as well. Since Model 2 is more simple but performs just as well as the full model, it might be the best choice.


```{r}
# Model 2
model2 <- lm(turnout ~ income + pct_bachelors_higher + pct_unemployed + median_age + pct_white, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ income + pct_bachelors_higher + pct_unemployed + median_age, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ population + income + pct_bachelors_higher + pct_unemployed + median_age + pct_white + income:pct_bachelors_higher, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

This analysis helps show what kinds of places have more people who vote, like areas with higher income, more education, or older residents. It can help reporters explain why some communities turn out to vote more than others. You could use these findings to highlight which factors seem to matter most and tell real stories about how they affect voters.
