---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The first thing I noticed was how many people are making six figures. Initially I thought the majority of employees with "senior" or "deputy" in their title made six figures but as I continued to examine the data I noticed that was not always the case. I noticed that employees withe "president" in their title were more consistently making more money, understandably so. There is a good amount of range within the salaries some making five figures but even beyond that there is a good spread among those making six. I will be interested to see the median of this data. I noticed outliers with detailees salaries listed as "0," this will definitely skew the mean of the data. Not all detailees have a salary of "0" though so I am curious why.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(salary))
wh_salaries |> summarize(median_salary = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary = mean(salary))
wh_salaries |> summarize(median_salary = median(salary))
wh_salaries |> summarize(min_salary = min(salary))
wh_salaries |> summarize(max_salary = max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The median beng greater than the mean lets me know that the data is skewed to the left. This confirms that the outliers with a "0" listed as a salary pulled the mean negatively. I would have liked to see the data without any outliers to gauge a more accurate mean.The minimum and maximum values don't really confirm the range that I anticipate because of the outlier. I think we could calculate the IQR to get a more complete pcture of Whte House Salaries. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I noticed several high-ranking officials earning less than what I expected while others with vague titles make significantly more. I want to know why this is the case how salaries are determined. I am immediately drawn to the unexplained outliers. I think there is a good story to figure out why people are listed as White House employees yet they aren't paid as such. The first thing I would do is find all the detailees and then determine who has a salary and whose salary is "0." I'd then look further into the job descriptions to try and figure out the salary disparities within positions. I would ttry and see if they were compensated in other ways.
