---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities_crime <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities_crime|> mutate (violent_rate_2019 = violent2019/pop2019 * 10000, property_rate_2019 = property2019/pop2019 * 10000, violent_rate_2020 = violent2020/pop2020 * 10000, property_rate_2020 = property2020/pop2020 * 10000) 
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

I noticed the violent rate had a lot more zeros amongst the counties in 2020 however the max was signicantly higher in 2020. I would say the violent crime rate went up. The property rate also saw more zeros but of of first glance it seems the rate was not as high. Ocean City had the highest rates amongst both categories. To me the story is Maryland crime is getting better but I would not say at drastic rates. The story is that Maryland is generally safe but some counties are worse off than others.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate (property_pct_change = (`property_rate_2020` - `property_rate_2019`)/`property_rate_2019` * 100, violent_pct_change = (`violent_rate_2020` - `violent_rate_2019`)/`violent_rate_2019` * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The property rate percentage changes looked consistent at first but then I found an outlier with Crisfield having over a 300% change. I immediately want to know why this number doubled. There were significant increases n the violent percent changes with the highest one being 895%. To me the story is Maryland is generally experiencing less property and violent crime rates however, certaiin areas are seeing a significant increase. My story angle would be to figure out why.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Baltimore was seen as one of the most dangerous cities in Maryland in 2019 but both the violent and property crime rates saw a decrease in 2020.

While Baltimore still had the second highest violent crime rate,the city saw a decrease by almost 13% and almost 26% for the property crime rate. It is worth noting that the city still saw significantly decreased crime rates despite reporting a larger population. 

Mayor Brandon Scott ran on the platform of decreased violence and his policies seemed to be effective throughout his first year in office.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Take the HW4_MD_CityCrime.Rmd notebook you worked on and add a section at the bottom called Task 10. Your task will be to take the violent crime rates for 2020 and do three things:

Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?
Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.
Do the results change your best story idea from the previous assignment? Why or why not?
As usual, save the notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

```{r}
md_cities_percentiles <- md_cities_change <- md_cities_rates |> mutate (percentile = ntile(md_cities_rates, 100))
```

10 cities are above the 90th percentile

```{r}
md_cities_percentiles |> summarize(violent_crime_IQR = IQR(violent_rate_2020, na.rm = TRUE))
```
This does change my idea for the best story but not subject wise, it just changes my angle. 50% of Maryland cities have a violent crime rate of about 42%, Baltimore's dropped however it dropped to 160% placing the city in one of the lowest percentiles. While originally I wanted my story to be on Baltimore's reduction in violent crime I would now mention that it's safer but it still remains extremely dangerous. 

