---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
m_m <- read_csv ("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset

```{r}
sample_5 <- m_m |> 
  sample_n(5)
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms
i calculated the mean red MandMs and sd of the sample
```{r}
sample_5 |> summarize(mean = mean(red), sd = sd(red))
```


### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4
```{r}
m_m |> summarize(mean = mean(red), sd = sd(red))
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The means are not similar and I think this is a result of the smaller sample size. The standard deviations don't vary as much only differing by about 1 number.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?
The data shows me that MandMs are very inconsistent and their data is spread out. As a result, random sampling will also vary greatly. 100 is already a small population which could explain why the data is hard to make inferences for.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
