---
title: "Funding Disparities Between HBCUs and PWIs"
author: "Nyla Cherry"
---

# Final Project Notebook

## Introduction

This project investigates funding disparities between Historically Black Colleges and Universities (HBCUs) and Predominantly White Institutions (PWIs). The primary hypothesis is that HBCUs receive significantly less federal and state funding per full-time equivalent (FTE) student than comparable PWIs, contributing to gaps in institutional resources, graduation outcomes, and academic support services. 

I am using data from 2023 and 2024 to track funding trends over time and analyze how disparities correlate with institutional outcomes such as spending, graduation rates, and access to academic resources.

# Data Sources

I am using two main sources:

1. **IPEDS (Integrated Postsecondary Education Data System) - U.S. Department of Education**
   - Source: [IPEDS Data Center](https://nces.ed.gov/ipeds/datacenter/InstitutionByName.aspx?sid=4feaf0bb-0f65-49c0-b588-2580c07d593d&rtid=1)
   - IPEDS provides downloadable CSVs with data on funding, enrollment, institutional characteristics, and outcomes.
   - After using College Navigator to select my 20 institutions, I then manually entered their names into "select institutions". For 2023 and 2024 I selected 12 variables: 
1. Under Finance
- Federal appropriations and grants

- State appropriations

- Total revenues

- Academic support expenditures

- Total core expenses

2. Under Fall Enrollment
- Full-time equivalent (FTE) enrollment

3. Under Institutional Characteristics
- Name

- Unit ID

- State

- Control (public)

- Sector
   

2. **NCES College Navigator**
  - Source: [College Navigator](https://nces.ed.gov/collegenavigator/)
  - This site is used for supplementary institution-level data (e.g., classification as HBCU, tuition, size).
  - I chose 10 4-year public HBCUs that varied in both location and size. I then chose comparable PWIs for each HBCU. I focused on 4-year public PWIs that were close in enrollment and location to my already selected HBCUs to try to keep the comparison as fair as possible. 


## Data Loading

# Data Set Up
```{r setup}
library(tidyverse)
library(readr)
library(janitor)
```

# Load 2023 Data
```{r}
year1_data <- read_csv("/Users/lornizzle/Downloads/jour405 final proj data/2023data.csv")
```

# Load 2024 Data
```{r}
year2_data <- read_csv("/Users/lornizzle/Downloads/jour405 final proj data/2024data.csv")
```
### Current Progress

The dataset includes institutional-level data for 10 public HBCUs and 10 matched public PWIs from the 2023 and 2024 academic years, with each institution identified by UnitID. It contains variables on federal and state appropriations, instructional spending, total core expenses, academic support, and full-time equivalent (FTE) enrollment, which are critical for calculating per-student funding and expenditures. Initial observations show that PWIs generally receive higher appropriations per FTE student and allocate more to instruction and academic resources than HBCUs, supporting the hypothesis of funding-related disparities. However, the dataset lacks student outcome metrics like graduation rates by race or income, which could further clarify the impact of funding gaps. Additional data on financial aid, faculty ratios, or detailed Carnegie classifications may be needed to strengthen comparisons, and it would be helpful to confirm whether any spending categories overlap to ensure accurate analysis.

##Tests I Could Run
Based on the analyses we've practiced in class, I plan to use several methods to examine funding disparities between HBCUs and PWIs. First, I will calculate key descriptive statistics (mean, standard deviation, minimum, and maximum) for variables like appropriations per FTE, instructional spending, and core expenses. To test for statistically significant differences between HBCUs and PWIs, I will use ANOVA, followed by Tukey’s HSD test if I expand the comparison to more than two institution types or regions. If appropriate, I may also apply linear regression models to explore whether funding levels predict instructional spending or outcomes like graduation rates. For any categorical comparisons, such as regional classification or control type, I may use a chi-square test to assess differences in distributions. Additionally, I’ll compute percent changes in funding or spending between 2023 and 2024 to examine short-term shifts across both groups.

##Trials and Error
# Merge datasets
```{r}
merged_data <- merge(year1_data, year2_data, by = `institution name...2`, suffixes = c("_2023", "_2024"))
```

I got this code from google, how do I properly merge the datasets? I was only able to download the data by separate years and I want to combine the data sets. This is saying the column isnt found but i know it is in the data set because I used it for other calculations.
#1. Descriptive statistics
```{r}
descriptive_stats23 <- year1_data |> 
  summarise(
    mean = mean(`F2223_F1A.Pell grants (federal)`),
    sd = sd(`F2223_F1A.Pell grants (federal)`),
    n = n(),
    min = min(`F2223_F1A.Pell grants (federal)`),
    max = max(`F2223_F1A.Pell grants (federal)`)
  )
```
```{r}
descriptive_stats24 <- year2_data |> 
  summarise(
    mean = mean(`F2122_F1A_RV.Pell grants (federal)`),
    sd = sd(`F2122_F1A_RV.Pell grants (federal)`),
    n = n(),
    min = min(`F2122_F1A_RV.Pell grants (federal)`),
    max = max(`F2122_F1A_RV.Pell grants (federal)`)
  )
```
#2. Finding St Appropriations per FTE
```{r}
year1_data <- year1_data |>
  mutate(State_Appropriations_per_FTE = `F2223_F1A.State appropriations` / `DRVEF2023.Full-time equivalent fall enrollment`)

year2_data <- year2_data |>
  mutate(State_Appropriations_per_FTE = `F2122_F1A_RV.State appropriations` / `DRVEF2022_RV.Full-time equivalent fall enrollment`)
```
#3. Find HBCUs
```{r}
hbcu_list <- c("Alabama A & M University", "Albany State University", "Alcorn State University", "Bowie State University","Delaware State University","Florida A&M University","Jackson State University","North Carolina A & T State University","Prairie View A & M University","Virginia State University")

# Add Institution_Type column
year1_data <- year1_data |>
  rename(Institution = `institution name...2`) |>
 mutate(Institution_Type = ifelse(Institution %in% hbcu_list, "HBCU", "PWI"))
```

```{r}
hbcu_list <- c("Alabama A & M University", "Albany State University", "Alcorn State University", "Bowie State University","Delaware State University","Florida A&M University","Jackson State University","North Carolina A & T State University","Prairie View A & M University","Virginia State University")

# Add Institution_Type column
year2_data <- year2_data |>
  rename(Institution = `institution name...2`) |>
 mutate(Institution_Type = ifelse(Institution %in% hbcu_list, "HBCU", "PWI"))
```

#4. ANOVA Tests
```{r}
anova_result <- aov(State_Appropriations_per_FTE ~ Institution_Type, data = year1_data)
summary(anova_result)

anova_result <- aov(State_Appropriations_per_FTE ~ Institution_Type, data = year2_data)
summary(anova_result)
```
#ANOVA Results
In 2023, the average state appropriations per FTE across the 20 institutions was about $16.9 million, with a standard deviation of $9.8 million, indicating substantial variation in funding levels. An ANOVA test comparing HBCUs and PWIs showed no statistically significant difference in appropriations per FTE (p = 0.435), suggesting that any observed funding disparities between the groups could be due to random variation.

In 2024, the average state appropriations per FTE among the 20 institutions was approximately $15.8 million, with a standard deviation of $9.4 million, again showing wide variation in funding. The ANOVA results indicated no statistically significant difference between HBCUs and PWIs in appropriations per FTE (p = 0.402), suggesting funding levels remained relatively comparable across institution types.

Between 2023 and 2024, the average state appropriations per FTE decreased slightly from $16.9 million to $15.8 million, with variability remaining high in both years. ANOVA results for both years showed no statistically significant differences in funding between HBCUs and PWIs, indicating consistent patterns across time.
